---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# TimeSeriesAnalysis

<!-- badges: start -->
<!-- badges: end -->

TiSA: TimeSeriesAnalysis - Is a transcriptomic analysis tool for both RNA sequencing and microarray data

## Overview

TimeSeriesAnalysis (TiSA) is an analysis and visualization package for RNAseq and
microarray data. TS extracts significant genes from time course
transcriptomic data by performing differential gene expression on both
the conditional and temporal axes. It then employs partitioning
algorithm based on recursive thresholding (PART) clustering to identify
small genomic clusters of relevance, followed by running the clusters
through gprofiler to reveal the biological relevance of each cluster.

## TS performs: 

* Data normalization and processing 
* PCA plots 
* Differential gene expression (conditional and temporal) 
* PART clustering 
* Heatmaps for both differential expression summary and clustering results
* Trajectory of identified clusters 
* Gprofiler (functional enrichment) analysis of clusters 
* Dotplots and MDS plots of Gprofiler results 
* Nearest ancestor clustering of GOs 
* GO ancestor queries


## Installation

You can install the stable version of TimeSeriesAnalysis like so:
``` r
install.packages("devtools")
devtools::install_github("Ylefol/TimeSeriesAnalysis@master")
```
The development version can be installed by replacing 'master' with 'dev'.

Devtools is required for the installation of this package. To install Devtools, visit the 
[Devtools main page](https://www.r-project.org/nosvn/pandoc/devtools.html)

Certain bioconductor packages will have to be installed before
installation of TimeSeriesAnalysis. The code snippet below gives the method along with
some of the packages that should be installed via this method.

``` r
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
bio_pkgs <- c('DESeq2','GOSemSim','GO.db','limma','ComplexHeatmap',
'AnnotationDbi','GenomeInfoDb','GenomicRanges','Biobase','S4Vectors',
'BioCGenerics','MatrixGenerics','IRanges','BiocFileCache'
'SummarizedExperiment','org.Hs.eg.db','org.Mm.eg.db','org.Ce.eg.db')
BiocManager::install(bio_pkgs)
```

## Rmarkdown format

TimeSeriesAnalysis was developed with user-friendliness in mind, the
core script of the package is a Rmarkdown file with explanations of what
each code block performs. Computationally intensive code blocks save
their work in order to avoid loss of computation time.
TimeSeriesAnalysis comes with a test dataset. Users are recommended to
first run the package with the test data to then modify the Rmarkdown
script with the necessary information for their purposes (input files,
organism, genes of interest etc…). The rmarkdown script and the expected
result from an example run can be found and downloaded from this
repository, they are located in the ‘rmarkdown_method’ folder. If users
prefer a script approach, the equivalent of the rmarkdown report is
provided in two scripts, one for the computation tasks and a second for
the analyses. These two scripts are found in the ‘script_method’ folder.

## Data

Three different datasets have been used to test this pipeline, one of which has been
preserved as an example datasets for the vignettes of this pipeline. More information
on the other datasets can be viewed through the publications listed below.

The PBMC dataset is a time series experiment with three time points that explores 
and compares three AICDA/AID (Activation Induced Cytidine DeAminase) activation cocktails.
The experiment seeked to identify which activation cocktail properly activated AID
through both the expression of the AID gene and the activation of class switch recombination.

The MURINE dataset is another AID related time series experiment, where mice were generated and
one was given an AID activation cocktail. The experiment contains one replicate per condition and
10 timepoints. Due to only having one replicate per condition per timepoint, subsequent timepoints
were combined to create an intermediary timepoint with two replicates in each condition.

The Celegans dataset is an ageing related experiment. Two celegans models were used: BY372 and N2.
The first being a parkinsons model and the second being a wild type. Each model was split into
two groups, one that would be fed krill oil and the other not (control). Each experiment
has three timepoints: Days 1, 3, and 6.

## Tutorial

A tutorial using the PBMC data can be found within the [pipelines website](https://ylefol.github.io/TimeSeriesAnalysis/). 
Additionally the documentation can be found in the references tab.

### Set-up of a sample file
In order for the pipeline to run properly, a sample file must be provided. Certain requirements  ust be met.
Column names should be as indicated below, all spaces should be avoided, timepoints should be numerical. It is also
important that the sample column contains the same names as the count files provided. Lastly, replicates
should always be followed by a '_#' as this numbering will be utilized for the organization of the replicates within 
certain visualizations. Note that in the below example the three files have the same name (ex: C_1). That is to represent
the 'C_1' sample at its three different time points.

sample file <- tibble::tribble(
  ~sample,    ~group, ~replicate, ~timepoint,
   "s-37", "control",      "C_1",          1,
   "s-38", "control",      "C_1",          3,
   "s-39", "control",      "C_1",          9,
   "s-52",     "IgM",    "IgM_1",          1,
   "s-53",     "IgM",    "IgM_1",          3,
   "s-54",     "IgM",    "IgM_1",          9,
   "s-77", "control",      "C_2",          1,
   "s-78", "control",      "C_2",          3,
   "s-79", "control",      "C_2",          9,
   "s-92",     "IgM",    "IgM_2",          1,
   "s-93",     "IgM",    "IgM_2",          3,
   "s-94",     "IgM",    "IgM_2",          9
  )

require(knitr)
kable(sample file, digits = 3, row.names = FALSE, align = "c",
              caption = NULL)

## Microarray based data

For microarray data, a streamlined method is in the works. Currently microarray data needs to be inputed 
as a Elist, specifically an E list saved as a rds object.
``` r
my_path_data<-'data/micro_arr/my_limma_dta.rds'
my_path_sample_dta<-'data/micro_arr/sample_file.csv'

#Set-up time series object parameters
diff_exp_type<-'limma' 
```
The rds file given to 'my_path_data' contains the Elist produced by limma processing
of microarray data. It is also important to set the differential gene expression type (diff_exp_type)
to 'limma'. 

## Using the pipeline

NOTE: You must still install the TimeSeriesAnalysis package as defined above

Since the pipeline's source code is available, individuals are free to take the code
and adapt it as needed. However for users who would like the easiest approach to use the
pipeline as is, it is recommended to clone or download this github repository to your local computer
and work from the Rmarkdown method folder within the repository.
Cloning or downloading can be done from the [TimeSeriesAnalysis github page](https://github.com/Ylefol/TimeSeriesAnalysis)
by using the green button called 'code'.
From there, the Rmarkdown file can be edited.

To edit the file, it is recommended to use rstudio, which can be downloaded [here](https://www.rstudio.com/products/rstudio/download/).
Note that it is recommended to download the free version of RStudio Desktop, not RStudio Server.

The only code block of code is the parameter set-up block (2nd) as well as the title
at the very top of the Rmarkdown document.

## Launching the pipeline

The pipeline can be launched from rstudio itself by using the 'knit' button, or it can be launched from a command line using the
following command from the TimeSeriesAnalysis repository:
```{r, engine = 'bash', eval = FALSE}
Rscript -e "rmarkdown::render('rmarkdown_method/TS_analysis.Rmd',output_file='TS_analysis.html')"
```

## Publication

A manuscript detailing the TimeSeriesAnalysis pipeline published with NAR genomics and bioinformatics.
[TiSA: TimeSeriesAnalysis—a pipeline for the analysis of longitudinal transcriptomics data](https://academic.oup.com/nargab/article/5/1/lqad020/7069286)
The version/DOI of this pipeline used within the manuscript is 10.5281/zenodo.7616032.

A ageing related study was performed in part through the use of TiSA.
[Krill oil protects dopaminergic neurons from age-related degeneration through temporal transcriptome rewiring and suppression of several hallmarks of aging](https://www.aging-us.com/article/204375/text)

A longitudinal COVID-19 study was performed using TiSA.
[Longitudinal whole blood transcriptomic analysis characterizes neutrophil activation and interferon signaling in moderate and severe COVID-19](https://www.nature.com/articles/s41598-023-37606-y)

## Additional information
The 'clusterGenomics' package was no longer maintained on CRAN and therefore the necessary
scripts were brought over to this package for it's implementation without the need
to download the clusterGenomics package from source.
Within this package, in the 'R/clusterGenomics.R' code was written by the authors of 
clusterGenomics. It was imported to TimeSeriesAnalysis by the authors of
TimeSeriesAnalysis.
